Spatio-angular fluorescence microscopy\\ II. Paraxial 4f imaging

### 2018-12-22 papers/basic-theory/paper/paper.tex
Patrick's split edits.

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We model an aplanatic microscope imaging an ensemble of in-focus
  fluorescent dipoles as a linear Hilbert-space operator, and we show that the
  operator takes a particularly convenient form when expressed in a basis of
  complex exponentials and spherical harmonics---a form we call the dipole
  spatio-angular transfer function. We discuss the implications of our analysis
  for all quantitative fluorescence microscopy studies and lay out a path
  towards a complete theory. % We demonstrate our formalism by
  % analyzing a single-view fluorescence microscope without using the monopole or
  % scalar approximations. We show that this imaging system has a spatio-angular
  % band limit, and we exploit the band limit to perform efficient simulations.
  % Notably, we show that information about the out-of-plane orientation of
  % ensembles of in-focus fluorophores is recorded by paraxial fluorescence
  % microscopes.
\end{abstract*}

>>>>

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy, providing a unified framework for analyzing systems that image
  single molecules or ensembles of molecules labeled with fluorescent dipoles.
  We model an aplanatic microscope imaging an ensemble of in-focus fluorescent
  dipoles as a linear Hilbert-space operator, and we show that the operator
  takes a particularly convenient form when expressed in a basis of complex
  exponentials and spherical harmonics---a form we call the dipole
  spatio-angular transfer function. We discuss the implications of our analysis
  for all quantitative fluorescence microscopy studies and lay out a path
  towards a complete theory. % We demonstrate our formalism by
  % analyzing a single-view fluorescence microscope without using the monopole or
  % scalar approximations. We show that this imaging system has a spatio-angular
  % band limit, and we exploit the band limit to perform efficient simulations.
  % Notably, we show that information about the out-of-plane orientation of
  % ensembles of in-focus fluorophores is recorded by paraxial fluorescence
  % microscopes.
\end{abstract*}

### 2018-12-22 papers/para-4f/paper/paper.tex
Patrick's split edits.

% Talon's custom libraries
\usepackage{empheq}
\usepackage{bm}
\usepackage[normalem]{ulem}

@@

\begin{abstract*}
  We analyze a single-view fluorescence microscope in a $4f$ geometry
    without using the monopole or scalar approximations. We show that this
    imaging system has a spatio-angular band limit, and we exploit the band
    limit to perform efficient simulations. Notably, we show that information
    about the out-of-plane orientation of ensembles of in-focus fluorophores is
    recorded by paraxial fluorescence microscopes. Additionally, we show that
    deconvolution algorithms that use the monopole approximation may make biased
    estimates of fluorophore concentrations, but these biases are small when the
    sample contains either many randomly oriented fluorophores in each
    resolvable volume or unconstrained rotating fluorophores.
\end{abstract*}

@@

We begin in section \ref{sec:theory} by specifying the imaging geometry and
  defining \textit{pupil functions} for both monopole and dipole imaging
  systems. We explicitly relate the pupil functions to the coherent transfer
  functions to establish a connection between physical calculations and the
  transfer functions. Next, in section \ref{sec:results} we calculate the
  monopole and dipole transfer functions in closed form, and we use these
  transfer functions to perform efficient simulations with four numerical
  phantoms. Finally, in section \ref{sec:discussion} we discuss the results and
  expand on how the pupil functions can be used to develop improved models for
  spatio-angular microscopes.

@@

@@

 \section{Discussion}\label{sec:discussion}
 \subsection{Comparing monopole and dipole models}
 The only case when the dipole and monopole transfer functions match
   exactly is when the sample consists of dipoles that are completely
   constrained to the transverse plane of a paraxial imaging system. Applying
   the monopole approximation in any other situation can lead to biased
   estimates of the fluorophore concentrations.

 However, the common justifications for the monopole approximation---the
   fluorophores are rotationally unconstrained or there are many randomly
   oriented fluorophores in a resolvable volume---are good justifications in all
   but the highest SNR regimes. The effects of the dipole model become apparent
   in lower SNR regimes as the rotational constraints on the dipoles increase
   (assuming there are out-of-plane dipole components).

>>>>

% Talon's custom libraries
\usepackage{empheq}
\usepackage{bm}
\usepackage[normalem]{ulem}
\usepackage{booktabs}

@@

\begin{abstract*}
  We investigate the properties of a single-view fluorescence microscope
  in a $4f$ geometry when imaging fluorescent dipoles without using the
  monopole or scalar approximations. We show that this imaging system has a
  spatio-angular band limit, and we exploit the band limit to perform efficient
  simulations. Notably, we show that information about the out-of-plane
  orientation of ensembles of in-focus fluorophores is recorded by paraxial
  fluorescence microscopes. Additionally, we show that the monopole
    approximation may cause biased estimates of fluorophore concentrations, but
  these biases are small when the sample contains either many randomly oriented
  fluorophores in each resolvable volume or unconstrained rotating fluorophores.
\end{abstract*}

@@

We begin in section \ref{sec:theory} by specifying the imaging geometry and
defining \textit{pupil functions} for imaging systems with and without the
monopole approximation. We explicitly relate the pupil functions to the coherent
transfer functions to establish a connection between physical calculations and
the transfer functions. Next, in section \ref{sec:results} we calculate the
monopole and dipole transfer functions in closed form, and we use these transfer
functions to perform efficient simulations with four numerical phantoms.
Finally, in section \ref{sec:discussion} we discuss the results and expand on
how the pupil functions can be used to develop improved models for
spatio-angular microscopes.

@@

After expressing the operators in a delta function basis we explored the
  form of the operators with several other choices of basis functions. Tables 1
  and 2 summarize our results.

  \begin{table}[t]
    \centering
  \begin{tabular}{lll}
    \toprule
    Quantity&Symbol&Relationships\\
    \midrule
    Monopole density&$f(\ro)$&---\\
    Monopole spectrum&$F(\bv)$&$=\mc{F}_{\mbb{R}^2}\left\{f(\ro)\right\}$\\
    \midrule
    Monopole coherent spread function&$c(\rd - \ro)$&---\\
    Monopole coherent transfer function&$C(\taup)$&$=\mc{F}_{\mbb{R}^2}\left\{c(\rd - \ro)\right\}$\\
    Monopole point spread function&$h(\rd - \ro)$&$=|c(\rd - \ro)|^2$\\
    Monopole transfer function&$H(\bv)$&$=\mc{F}_{\mbb{R}^2}\left\{h(\rd-\ro)\right\}$\\
    &&$= \int_{\mbb{R}^2}d\taup\,C(\taup)C^{*}(\taup - \bv)$\\
    \midrule
    Scaled irradiance&$g(\rd)$&$=\int_{\mbb{R}^2}d\ro\, h(\rd - \ro)f(\ro)$\\
    Scaled irradiance spectrum&$G(\bv)$&$=\mc{F}_{\mbb{R}^2}\left\{g(\rd)\right\}=H(\bv)F(\bv)$\\
    \bottomrule
\end{tabular}
\caption{Summary of relevant quantities in fluorescence microscopy under the
  monopole approximation---see \cite{chandler2018} for derivations.
  $\mc{F}_{\mbb{R}^2}$ denotes a two-dimensional Fourier transform.}
\end{table}

  \begin{table}[t]
    \hspace{-1.5em}
  \begin{tabular}{lll}
    \toprule
    Quantity&Symbol&Relationships\\
    \midrule
    Dipole density&$f(\ro,\so)$&---\\
    Dipole spatial spectrum&$F(\bv,\so)$&$=\mc{F}_{\mbb{R}^2}\left\{f(\ro,\so)\right\}$\\
    Dipole angular spectrum&$F_{\ell}^m(\ro)$&$=\mc{F}_{\mbb{S}^2}\left\{f(\ro,\so)\right\}$\\
    Dipole spatio-angular spectrum&$\msf{F}_{\ell}^m(\bv)$&$=\mc{F}_{\mbb{R}^2}\left\{F_{\ell}^m(\ro)\right\} = \mc{F}_{\mbb{S}^2}\left\{F(\bv, \so)\right\}$\\
    \midrule
    Dipole coherent spread function&$\mb{c}(\rd - \ro, \so)$&---\\
    Dipole coherent transfer function&$\mb{C}(\taup,\so)$&$=\mc{F}_{\mbb{R}^2}\left\{\mb{c}(\rd - \ro, \so)\right\}$\\
    Dipole point spread function&$h(\rd - \ro,\so)$&$=|\mb{c}(\rd - \ro, \so)|^2$\\
    Dipole spatial transfer function&$H(\bv,\so)$&$=\mc{F}_{\mbb{R}^2}\left\{h(\rd-\ro,\so)\right\}$\\
    &&$= \int_{\mbb{R}^2}d\taup\,\mb{C}(\taup,\so)\mb{C}^{\dagger}(\taup - \bv,\so)$\\
    Dipole angular transfer function&$H_{\ell}^m(\rd - \ro)$&$=\mc{F}_{\mbb{S}^2}\left\{h(\rd-\ro,\so)\right\}$\\
    Dipole spatio-angular transfer function&$\msf{H}_{\ell}^m(\bv)$&$=\mc{F}_{\mbb{R}^2}\left\{H_{\ell}^m(\rd - \ro)\right\} = \mc{F}_{\mbb{S}^2}\left\{H(\bv, \so)\right\}$\\
    \midrule
    Scaled irradiance&$g(\rd)$&$=\int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h(\rd - \ro, \so)f(\ro,\so)$\\
    &&$= \sum_{\ell m}\int_{\mbb{R}^2}d\ro\,H_{\ell}^m(\rd - \ro)F_{\ell}^m(\ro)$\\
    Scaled irradiance spectrum&$G(\bv)$&$=\mc{F}_{\mbb{R}^2}\left\{g(\rd)\right\}$\\
            &&$=\int_{\mbb{S}^2}d\so\, H(\bv,\so)F(\bv, \so)$\\
            &&$=\sum_{\ell m} \msf{H}_{\ell}^m(\bv) \msf{F}_{\ell}^m(\bv)$\\
    \bottomrule
\end{tabular}
  \caption{Summary of relevant quantities in spatio-angular dipole imaging---see \cite{chandler2018} for derivations. $\mc{F}_{\mbb{R}^2}$ denotes a two-dimensional Fourier transform, and $\mc{F}_{\mbb{S}^2}$ denotes a spherical Fourier transform.}
\end{table}

@@

 \section{Discussion}\label{sec:discussion}
 \subsection{Comparing monopole and dipole models}
 The only case when the dipole and monopole transfer functions match exactly is
 when the sample consists of dipoles that are completely constrained to the
 transverse plane of a paraxial imaging system. Applying the monopole
 approximation in any other situation can lead to biased estimates of the
 fluorophore concentrations. To see how these biases manifest, consider the
   irradiance pattern created by an ensemble of dipoles oriented along the optic
   axis---see Figs. \ref{fig:ph1}, \ref{fig:ph2}, \ref{fig:ph3}, or \ref{fig:ph4}.
   Any reconstruction scheme that uses the monopole approximation would
   attribute the irradiance doughnut to a doughnut of monopoles instead of
   axially oriented dipoles, which is a clear example of a biased estimate caused
   by model mismatch.

   However, the common justifications for the monopole approximation---that the
   fluorophores are rotationally unconstrained or that
   there are many randomly
   oriented fluorophores in a resolvable volume---are good justifications in all
   but the highest SNR regimes. The effects of the dipole model become apparent
   in lower SNR regimes as the rotational constraints on the dipoles increase
   (assuming there are out-of-plane dipole components).
